---
title: "Forced Turnover:<br>Evaluating Pressing Effectiveness in Soccer"
author: Natalie Rayce, David Almona
editor: visual
format:
  revealjs:
    slide-number: true
    embed-resources: true
    preview-links: auto
    # transition: slide
    # smaller: true
---

## Introduction

::: fragment
**Pressing:** a defensive tactic where players apply coordinated pressure on the opponent with the ball to force mistakes, win back possession, and quickly transition to attack
:::

::: fragment
**Forced Turnover:** when a player loses possession due to opponent pressure, resulting in the opposing team gaining control. This includes misplaced passes, interceptions, successful tackles, or losing control under pressure - all direct results of effective defensive pressure
:::




## Main Question
Can the effectiveness of a press in soccer be predicted using factors such as spatial context, pressing dynamics, game context and situational factors?


## Data

::: fragment
-   **Dataset:** 521 matches in the MLS 2023 season
:::

::: fragment
-   **Source:** SkillCorner
:::

::: fragment
-   **Three data types:**
    -   **Match information:** game details (teams, pitch, referee)
    -   **Event data:** player actions (passes, shots, tackles) with timestamps and coordinates
    -   **Tracking data:** real-time positions of all players and ball (10 Hz)
:::

## Methods: Detecting "Pressure Zones"

::: fragment
-   Be within **3 meters** behind the ball carrier or **9 meters** in front of the ball carrier, **AND**
:::


::: fragment
-   Either be approaching at more than 1.0 meters per second OR show close engagement behaviors
:::

##

```{r}
library(magick)
#| label: fig-plot
#| include: TRUE
#| echo: FALSE
#| eval: TRUE
#| warning: FALSE
#| message: FALSE
#| fig.width: 24
#| fig.height: 16
#| fig.cap: "plot"

# save image as ggplot
my_plot_ggplot <- image_read("pressure_zone.png") %>% 
  image_ggplot(interpolate = TRUE)
my_plot_ggplot
```

## Methods: Detecting Pressure Sequences

::: fragment
-   Pressing actions were grouped into sequences if at least one defender continued pressing, allowing gaps of up to 1.5 seconds.
:::

::: fragment
-   A new sequence was defined after any pause longer than 1.5 seconds.
:::

::: fragment
-   For each sequence, the following features were recorded:
    -   duration, number of defenders involved and average defender velocity at the start
:::

## Feature Engineering

::: fragment
-   **Spatial Context:** Ball carrier position, distance to boundaries, field third
:::

::: fragment
-   **Pressing Dynamics:** Number of defenders, approach velocity, passing options
:::

::: fragment
-   **Game Context:** Score, game state (winning/losing/drawing), time remaining
:::

::: fragment
-   **Situational Factors:** How ball carrier gained possession (pass reception, interception, etc.), incoming pass characteristics (distance, height, range)
:::

## Modeling

::: fragment
- **Model:**  
  - Logistic Regression  
  - XGBoost  
  - All models evaluated using 10-fold cross-validation
:::

## Model Performance

```{r}
library(magick)
#| label: fig-plot
#| include: TRUE
#| echo: FALSE
#| eval: TRUE
#| warning: FALSE
#| message: FALSE
#| fig.width: 24
#| fig.height: 16
#| fig.cap: "plot"

# save image as ggplot
my_plot_ggplot <- image_read("model_summary.png") %>% 
  image_ggplot(interpolate = TRUE)
my_plot_ggplot
```

## Results

```{r}
library(magick)
#| label: fig-plot
#| include: TRUE
#| echo: FALSE
#| eval: TRUE
#| warning: FALSE
#| message: FALSE
#| fig.width: 24
#| fig.height: 16
#| fig.cap: "plot"

# save image as ggplot
my_plot_ggplot <- image_read("diverging_plot.png") %>% 
  image_ggplot(interpolate = TRUE)
my_plot_ggplot
```
##
```{r}
library(magick)
#| label: fig-plot
#| include: TRUE
#| echo: FALSE
#| eval: TRUE
#| warning: FALSE
#| message: FALSE
#| fig.width: 24
#| fig.height: 16
#| fig.cap: "plot"

# save image as ggplot
my_plot_ggplot <- image_read("volume_vs_effectiveness_plot.png") %>% 
  image_ggplot(interpolate = TRUE)
my_plot_ggplot
```

## Discussion

::: fragment
-   **Limitations:**
    -   Inaccuracy in SkillCorner data
    -   Individual player skill and tactical tendencies not considered in the model
    -   Our model does not account for pitch control
    
:::

::: fragment
-   **Future Works:**
    -   Incorporate pressing intensity calculations
    -   Incorporate pitch control
    
:::